### 2023-08-28 reports/paper/paper.qmd
Reduce abstract length

**Authors:** Kevin T. Chen^a^, Sally Picciotto^b^, Ellen A. Eisen^b^

@@

^b^ University of California, Berkeley School of Public Health, Division
of Environmental Health Sciences

**Corresponding author:**\
Kevin T. Chen\
2121 Berkeley Way\
Room 5302, Desk 5305-1\
Berkeley, CA 94720-7360\
kevchen\@berkeley.edu\
(510) 387-3327

**Running head suggestion:** Supportable interventions for reducing NHL
risk.

@@

**Sources of financial support:** This work was supported by Training
Grant T42OH008429 and Research Project Grant R01OH011092, both funded by
the National Institute for Occupational Safety and Health (NIOSH) /
Centers for Disease Control and Prevention (CDC).

@@

**Word count** **(Limit: 1500 to 4000 words):** 3812 (approx)

@@

**Methods:** We estimated counterfactual risk of NHL between 1985 and
2015 in the UAW-GM cohort of autoworkers at three Michigan plants (n =
33,134) under hypothetical supportable interventions on exposure to
soluble MWF. A supportable intervention reduces annual daily average
exposures above a specific target limit to that limit when supported by
data. Otherwise, the supportable intervention reduces exposure to the
highest level below the target limit or not at all, depending on
support.

**Results:** 339 NHL cases occurred over the 30-year follow-up period.
Stronger target limits on soluble MWF exposure resulted in monotonic
reductions in NHL risk. Setting the target exposure limit at 0.5
mg/m^3^, the NIOSH recommended exposure limit, would have prevented 112
(95% CI: 57, 181) cases.

**Conclusion:** More stringent limits on exposure to soluble MWF provide
stronger protections against NHL. We expect the effects of enforcing the
target exposure limits uniformly to have even stronger protective
effects. Strengthening exposure limits for MWF may protect workers from
NHL during the anticipated boom in domestic manufacturing.

@@

Non-Hodgkin Lymphoma (NHL) incidence in the United States doubled
between 1973 and 1994, before plateauing at around 19 per 100,000
persons per year, making it the seventh most common cancer in the
country [@seer_1994; @Ekstrom-Smedby_2006]. The strongest known risk
factor of NHL is immunosuppression, both congenital and acquired
[@Filipovich_1992; @Chiu_2015]. However, since immunodeficiency and
infection with HIV are rare, they cannot fully explain the historic rise
or present burden of NHL [@Shiels_2013]. The rise in NHL incidence
coincided with a period of rapid and extensive chemicalization in
industry, agriculture and warfare; environmental and occupational
exposures may play an important explanatory role in the epidemiology of
NHL [@Nelson_2005; @Romero_2021].

@@

One challenge in estimating the causal effects of occupational exposures
on worker health is the Healthy Worker Survivor Effect (HWSE), the
process by which healthier individuals remain at work where they
accumulate more exposure while those more susceptible to the deleterious
health effects of exposure leave work [@Arrighi_1994]. An early causal
inference method in statistics developed to estimate causal effects in
longitudinal observational studies where the HWSE or other forms of
time-varying confounding/selection bias affected by past exposure may be
operating is the parametric g-formula [@Robins_1986; @Taubman_2009;
@Richardson_2013]. A central requirement necessary for causal inference
from observational data is positivity (overlap) ie adequate variation in
the exposure of interest within strata formed by confounder and exposure
histories [@Petersen_2012].

Common approaches for addressing violations in positivity are
model-based smoothing when using outcome regression methods, weight
truncation when using propensity methods, or a combination of both.
Another approach is to only consider causal contrasts that are supported
by the observed data ie those involving hypothetical interventions with
propensity score strictly greater than zero. Here, we defined
hypothetical supportable interventions on exposure to soluble MWF in the
United Auto Workers-General Motors (UAW-GM) occupational cohort and
estimated their effects on NHL incidence between 1985 and 2015 relative
to no exposure intervention while guaranteeing positivity.

@@

We estimated the cumulative incidence of NHL between 1985 and 2015 under
supportable interventions based on selected target exposure limits on
annual average daily exposure to soluble MWF by applying the the
hazard-extended iterative conditional expectation (ICE) parametric
g-formula [@Wen_2020]. First, we estimated the expected number of NHL
cases that we would observe if there were no censoring and no limit on
exposure. Then, we contrasted this case count to that under supportable
interventions based on five hypothetical target exposure limits and no
censoring. The five target exposure limits were (1) 2.0, (2) 1.0, (3)
0.5, (4) 0.25, and (5) 0.05 mg/m^3^. The National Institute for
Occupational Safety and Health (NIOSH) recommended exposure limit (REL)
for time-weighted average total particulate mass (PM) composed of MWF is
0.5 mg/m^3^ [@niosh_1998].

For each target exposure limit, period of follow-up, and stratum defined
by unique combinations of confounder and exposure histories, we found a
supportable exposure limit, which was the maximum observed value of
exposure at or below the target exposure limit, if such a value exists.
If all of the observed values were above the target limit, the
supportable limit was simply the maximum observed value. The supportable
intervention rule then reduces exposures above the supportable exposure
limit to that limit, but allows exposures at or below the limit to vary
according to their observed distribution. Applying the supportable
intervention rule to the observed distribution of exposure produces the
intervention distribution that defines the corresponding stochastic
dynamic intervention with guaranteed positivity. We estimated the effect
of supportable intervention rules based on the selected target exposure
limits, expressed as stochastic dynamic interventions, using the
hazard-extended ICE parametric g-formula [@Wen_2020].

Figure 1 presents three example scenarios where the target exposure
limit is 0.25 mg/m^3^ in all cases, but the supportable exposure limit
differs depending on what the observed data supports. In Figure 1a, the
supportable exposure limit is equal to the target exposure limit as some
individuals with that particular set of potential confounder and
exposure histories were observed to have experienced exposure at that
level. In Figure 1b, the supportable exposure limit is below the target
exposure limit. In Figure 1c, there is no limit on exposure because no
individuals in that stratum experienced exposure below the target
exposure limit.

@@

We used data from the UAW-GM cohort study, which included all hourly
workers at three automobile manufacturing plants in Michigan who had
worked at least three years by 1985. Past papers provide detailed
descriptions of the cohort [@Eisen_1992; @Eisen_2001]. The large size of
the study population and rich time-varying, quantitative MWF exposure
data enable the study of a relatively rare cancer and evaluate realistic
interventions on MWF exposure in a longitudinal cohort setting. The
present study population (N = 33,134) was restricted to the autoworkers
who were at work in 1941 or not yet hired, missing no more than half of
their employment history, and still alive at the start of follow-up for
cancer incidence on January 1, 1941. Autoworkers in the study population
were followed until NHL diagnosis, death, December 31, 2014, or the
oldest observed age at death, whichever came earlier.

@@

Quantitative measurement of time-varying MWF exposure is a
distinguishing strength of the UAW-GM cohort study relative to other
environmental or occupational health studies. Exposure assessment was
based on direct air sampling as well as company records. Company
industrial hygienists collected several hundred personal and area
samples for total particulate mass (mg/m^3^) composed of MWF over many
decades. Research industrial hygienists collected additional air
sampling data when the cohort study was launched in the mid 1980s
[@Woskie_1994; @Woskie_1994a]. These additional data combined with
historical data and company records constituted a job-exposure matrix of
quantitative 8-hour time-weighted average daily exposure estimates to
soluble, straight, and synthetic MWF for each combination of job,
department, and plant over time [@Hallock_1994]. Workers' annual average
daily exposure to each MWF type was determined by combining this
job-exposure matrix with employment records, which recorded time-varying
job type, department, and plant for each employee from hire to
termination or 1995, whichever came sooner. For employment records that
were at least half complete, gaps in the record were interpolated by
carrying forward the last known job type. The exposure assessment is
described in detail elsewhere [@Hallock_1994; @Woskie_1994;
@Woskie_2003].

@@

We investigated supportable interventions that guarantee positivity:
every value of exposure that could be assigned under the our stochastic
dynamic intervention has propensity score strictly greater than zero. We
treated co-exposure to straight and synthetic MWF, employment status,
cumulative time off, age, duration of employment, sex (male/female),
race (Black/white/unknown), and plant (Plant 1/Plant 2/Plant 3) as
potential confounders. Exposure to MWF, employment status, time off, and
duration of employment were lagged 20 years. An overview of the general
steps of the estimation procedure are presented below. Note that we
refer to discrete hazards simply as hazards.

@@

The observed number of NHL cases over the 30-year follow-up period was
339. Table 2 presents estimates of the counterfactual number of cases,
number of cases averted, and cumulative incidence ratios contrasting
supportable interventions on exposure to soluble MWF based on different
target exposure limits and no censoring. Under an intervention
eliminating censoring, the estimated number of cases was 502 (95% CI:
439, 555). Interventions based on stronger target exposure limits on
soluble MWF resulted in monotonically stronger reductions in the
estimated cumulative incidence of NHL. Setting the target exposure limit
at the NIOSH REL 0.5 mg/m^3^ would have averted 112 (95% CI: 57, 181)
NHL cases.

@@

Correct model specification is a standard assumption in all parametric
and semi-parametric analyses. In causal analyses of longitudinal cohort
studies, ICE g-formula estimators are less common than the classic
parametric g-formula [@Keil_2014]. However, a major limitation of the
classic g-formula is the g-null paradox: the guaranteed misspecification
of parametric models resulting in the false rejection of the null
hypothesis when the null is true and when there is time-varying
confounding affected by past exposure [@Naimi_2015; @McGrath_2022]. As
with all ICE g-formula estimators, the estimator we applied is not
subject to the g-null paradox. Furthermore, since ICE g-formula
estimators require fewer parametric modeling requirements than the
classic parametric g-formula, correct model specification may be
achieved more readily.

>>>>

**Authors:** Kevin T. Chen\^a, b\^, Sally Picciotto^c^, Ellen A.
Eisen^c^

@@

^b^ University of California, Berkeley Department of Statistics

^c^ University of California, Berkeley School of Public Health, Division
of Environmental Health Sciences

**Corresponding author:**\
Kevin T. Chen\
2121 Berkeley Way\
Room 5302, Desk 5305-1\
Berkeley, CA 94720-7360\
kevchen\@berkeley.edu\
(510) 816-3408

**Running head suggestion:** Hypothetical interventions with guaranteed positivity

@@

**Sources of financial support:** This work was supported by Training
Grant T42OH008429 and Research Project Grant R01OH011092, both from
the National Institute for Occupational Safety and Health (NIOSH) /
Centers for Disease Control and Prevention (CDC).

@@

**Word count** **(Limit: 1500 to 4000 words):** 3842 (approx)

@@

**Methods:** We defined hypothetical supportable interventions that
reduce annual daily average exposures above a specific target limit to
that limit when supported by data. When the target limit is not
supported, the intervention reduces exposure to the nearest supported
level of exposure. We estimated counterfactual risk of NHL between 1985
and 2015 in the UAW-GM cohort of autoworkers at three Michigan plants (n
= 33,134) under hypothetical supportable interventions on exposure to
soluble MWF.

**Results:** 339 NHL cases occurred over the 30-year follow-up period.
Stronger target limits on soluble MWF exposure resulted in monotonic
reductions in NHL risk. Setting the target exposure limit at 0.5
mg/m^3^, the NIOSH recommended exposure limit, would have prevented 124
(95% CI: 66, 202) cases.

**Conclusion:** More stringent limits on exposure to soluble MWF provide
stronger protections against NHL. We expect that uniformly enforcing the
target exposure limits, regardless of data support, would have even
stronger protective effects. Strengthening exposure limits for MWF may
protect workers from NHL during the anticipated boom in domestic
manufacturing.

@@

Non-Hodgkin Lymphoma (NHL) incidence in the United States doubled
between 1973 and 1994 before plateauing at around 19 per 100,000 persons
per year, making it the seventh most common cancer in the country
[@seer_1994; @Ekstrom-Smedby_2006]. The strongest known risk factor of
NHL is immunosuppression, both congenital and acquired
[@Filipovich_1992; @Chiu_2015]. However, since immunodeficiency and
infection with HIV are rare, they cannot fully explain the historic rise
or present burden of NHL [@Shiels_2013]. The rise in NHL incidence
coincided with a period of rapid and extensive chemicalization in
industry, agriculture and warfare; environmental and occupational
exposures may play an important explanatory role in the epidemiology of
NHL [@Nelson_2005; @Romero_2021].

@@

One challenge in estimating the causal effects of occupational exposures
on worker health is the Healthy Worker Survivor Effect (HWSE), the
process by which healthier individuals remain at work where they
accumulate more exposure while those more susceptible to the deleterious
health effects of exposure leave work [@Arrighi_1994]. The parametric
g-formula is an early causal inference method in statistics developed to
estimate causal effects in longitudinal observational studies where the
HWSE or other forms of time-varying confounding/selection bias affected
by past exposure may be operating [@Robins_1986; @Taubman_2009;
@Richardson_2013]. A central requirement necessary for causal inference
from observational data is positivity (overlap), *i.e.,* adequate
variation in the exposure of interest within strata formed by confounder
and exposure histories [@Petersen_2012].

Common approaches for addressing violations in positivity are
model-based smoothing when using outcome regression methods, weight
truncation when using propensity methods, or a combination of both.
Another approach is to only consider causal contrasts that are supported
by the observed data, *i.e.,* those involving hypothetical interventions
with propensity score strictly greater than zero. Here, we defined
hypothetical supportable interventions on exposure to soluble MWF in the
United Auto Workers-General Motors (UAW-GM) occupational cohort and
estimated their effects on NHL incidence between 1985 and 2015 relative
to no exposure intervention while guaranteeing positivity.

@@

We estimated the cumulative incidence of NHL between 1985 and 2015 under
supportable interventions based on selected target exposure limits on
annual average daily exposure to soluble MWF by applying the
hazard-extended iterative conditional expectation (ICE) parametric
g-formula [@Wen_2020]. First, we estimated the expected number of NHL
cases that we would observe if there were no censoring and no limit on
exposure. Then, we contrasted this case count to that under supportable
interventions based on five hypothetical target exposure limits and no
censoring. The five target exposure limits were (1) 2.0, (2) 1.0, (3)
0.5, (4) 0.25, and (5) 0.05 mg/m^3^. The National Institute for
Occupational Safety and Health (NIOSH) recommended exposure limit (REL)
for time-weighted average total particulate mass (PM) composed of MWF is
0.5 mg/m^3^ [@niosh_1998].

For each target exposure limit, period of follow-up, and stratum defined
by unique combinations of confounder and exposure histories, we found a
supportable exposure limit, which was the maximum observed value of
exposure at or below the target exposure limit, if such a value exists.
If all of the observed values were above the target limit, the
supportable limit was the minimum observed value. The supportable
intervention rule then reduces exposures above the supportable exposure
limit to that limit, but allows exposures at or below the limit to vary
according to their observed distribution. Applying the supportable
intervention rule to the observed distribution of exposure produces the
intervention distribution that defines the corresponding stochastic
dynamic intervention with guaranteed positivity. We estimated the effect
of supportable intervention rules based on the selected target exposure
limits, expressed as stochastic dynamic interventions, using the
hazard-extended ICE parametric g-formula [@Wen_2020].

Figure 1 presents three example scenarios where the target exposure
limit is 0.25 mg/m^3^ in all cases, but the supportable exposure limit
differs depending on what the observed data supports. In Figure 1a, the
supportable exposure limit is equal to the target exposure limit as some
individuals with that particular set of potential confounder and
exposure histories were observed to have experienced exposure at that
level. In Figure 1b, the supportable exposure limit is below the target
exposure limit. In Figure 1c, the supportable exposure limit, the
minimum observed level of exposure in that stratum, is greater than the
target exposure limit.

@@

We used data from the UAW-GM cohort study, which included all hourly
workers at three automobile manufacturing plants in Michigan who had
worked at least three years by 1985. Past papers provide detailed
descriptions of the cohort [@Eisen_1992; @Eisen_2001]. The large size of
the study population and rich time-varying, quantitative MWF exposure
data enable the study of a relatively rare cancer and evaluation of
realistic interventions on MWF exposure in a longitudinal cohort
setting. The present study population (N = 33,134) was restricted to the
autoworkers who were at work in 1941 or not yet hired, missing no more
than half of their employment history, and still alive at the start of
follow-up for cancer incidence on January 1, 1985. Autoworkers in the
study population were followed until NHL diagnosis, death, December 31,
2014, or the oldest observed age at death, whichever came earlier.

@@

Quantitative measurement of time-varying MWF exposure is a
distinguishing strength of the UAW-GM cohort study relative to other
environmental or occupational health studies. Exposure assessment was
based on direct air sampling as well as company records. Company
industrial hygienists collected several hundred personal and area
samples for total particulate mass (mg/m^3^) composed of MWF over many
decades. Research industrial hygienists collected additional air
sampling data when the cohort study was launched in the mid 1980s
[@Woskie_1994; @Woskie_1994a]. These additional data combined with
historical data and company records were used to construct a
job-exposure matrix of quantitative 8-hour time-weighted average daily
exposure estimates to soluble, straight, and synthetic MWF for each
combination of job, department, and plant over time [@Hallock_1994].
Workers' annual average daily exposure to each MWF type was determined
by combining this job-exposure matrix with employment records, which
recorded time-varying job type, department, and plant for each employee
from hire to termination or 1995, whichever came sooner. For employment
records that were at least half complete, gaps in the record were
interpolated by carrying forward the last known job type. The exposure
assessment is described in detail elsewhere [@Hallock_1994;
@Woskie_1994; @Woskie_2003].

@@

We investigated supportable interventions that guarantee positivity:
every value of exposure that could be assigned under our stochastic
dynamic intervention has propensity score strictly greater than zero. We
treated co-exposure to straight and synthetic MWF, employment status,
cumulative time off, age, duration of employment, sex (male/female),
race (Black/white/unknown), and plant (Plant 1/Plant 2/Plant 3) as
potential confounders. Exposure to MWF, employment status, time off, and
duration of employment were lagged 20 years. An overview of the general
steps of the estimation procedure are presented below. Note that we
refer to discrete hazards simply as hazards.

@@

The observed number of NHL cases over the 30-year follow-up period was
339. Table 2 presents estimates of the counterfactual number of cases,
number of cases averted, and cumulative incidence ratios contrasting
supportable interventions on exposure to soluble MWF based on different
target exposure limits and no censoring. Under an intervention
eliminating censoring, the estimated number of cases was 502 (95% CI:
439, 555). Interventions based on stronger target exposure limits on
soluble MWF resulted in monotonically stronger reductions in the
estimated cumulative incidence of NHL. Setting the target exposure limit
at the NIOSH REL 0.5 mg/m^3^ would have averted 124 (95% CI: 66, 202)
NHL cases.

@@

Correct model specification is a standard assumption in all parametric
and semi-parametric analyses. In causal analyses of longitudinal cohort
studies, ICE g-formula estimators are less common than the classic
parametric g-formula [@Keil_2014]. However, a major limitation of the
classic g-formula is the g-null paradox: the guaranteed misspecification
of parametric models resulting in the false rejection of the null
hypothesis when the null is true and when there is time-varying
confounding affected by past exposure [@Naimi_2015; @McGrath_2022]. As
with all ICE g-formula estimators, the estimator we applied is not
subject to the g-null paradox. Furthermore, since ICE g-formula
estimators require fewer parametric modeling requirements than the
classic parametric g-formula, correct model specification may be
achieved (or approximated) more readily.
